How is experience transformed into memory?

### 2018-09-02 paper/boneyard.tex
working on discussion [WIP]

\textbf{JRM STOPPED HERE}

>>>>

% DISCUSSION
Prior work on neural responses during naturalistic experiences and recall has largely focused on identifying brain regions whose responses are reliably similar across individuals ~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17}. This allows one to identify which regions might be processing or representing the stimulus, or retrieving details of the experience during recall. Our approach is fundamentally different. We ask: which brain regions during video viewing match the video structure and how individual participants will recall the video later (in terms of the temporal correlations of their recall topic trajectory)? This latter approach highlights regions such as the medial prefrontal cortex (mPFC) and medial temporal lobes (MTL) that may respond in idiosyncratic ways across individuals, but that nonetheless play an important role in encoding experience into memory. In other work, the mPFC has been suggested to play a role in the representation of a ``schema'', or a network of prior knowledge that is consistent with the current task ~\citep{KestEtal12, GilbMarl17}. One interpretation for mPFC's involvement in our work is that mPFC represents participant-specific relationships with the content of the video, and its representation during scene viewing is related to how a participant later describes the scene. MTL activity patterns also matched participant's recall, which is consistent with a large body of literature highlighting the role of the MTL in episodic memory encoding ~\citep{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.

More broadly, these findings have strong implications for how we assess memory in other naturalistic contexts, such as the classroom or in a doctor's office.  Whereas academic tests often measure students' performance using metrics such as the proportion of correctly answered questions, our work suggests that this approach might ``miss the forrest for the trees''. We view ``true learning'' as understanding key concepts (i.e., understanding central themes in the learned content and how they relate) rather than regurgitating the greatest number of facts. In addition to educational contexts, our approach may provide unique metrics that can be used to assist in the diagnosis of memory disorders, and other psychiatric disorders that influence memory. For example, while the quantity of information recalled could be roughly matched between a healthy and patient population, and other aspects of the memory (such as the shape, serial order, precision or distinctiveness) might be different. Thus, this work serves as a foundation for more nuanced approaches to memory assessment that consider the trajectory and specific contents of memory for a naturalistic experience.

Studying human memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. While this approach has advanced our understanding of human memory immensely, it does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience.

Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words. Our work characterizes recollection of an experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the video participants viewed by measuring the match between the shapes of the video's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory -- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape. In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g., percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

An important question for future work concerns the factors that drive an individual to sample their recall trajectory finely or coarsely. For example, given a short recall interval, would participants intuitively gravitate towards coarser samplings that still outline the basic shape of the video's topic trajectory? Or if participants were told that their narrations would be played back to other participants ~\citep{ZadbEtal17}, would that change the resolution or shape of their recalls? And over successive recountings of the same sequence of events or with more elapsed time between encoding and recall, how do the shapes of the trajectories change? For example, loss of detail would result in a "smoothing out" of the trajectories with each new retelling.

While we view this work as a major advance in characterizing and understanding human memory, as with any approach there are limitations. First, the approach relies on having a ``good'' model of the stimulus (e.g., one that describes the video in the same way as participants recall it), which is currently only achievable by a human hand-annotating each moment of the video.  To increase the scalability of this approach, future work could explore automated methods for extracting meaning from videos ~\citep{YuEtal16}.  Another potential limitation is that by its nature, the model extracts the ``gist'' of scenes from the video.  This provides desirable flexibility (e.g., participants can use different combinations of words to describe the scenes), but this comes at the expense of capturing specific details.  Thus, a future direction of this work will be to increase sensitivity to such details while maintaining flexibility.

To conclude, we'll revisit the question of what it means to remember something. Our view is that ``successful remembering'' is about recovering the ``trajectory'' of an experience, rather than the ability to recognize/recall any of its particular isolated details. Decades of research suggest that episodic memories are not veridical and context-free snapshots of the past, and so treating (and modelling) them as such is overly simplistic at best (and counterproductive at worst). While it's undeniable that these models have been useful in advancing our understanding of human memory, they are severely limited in their ability to explain memory for real life experiences. Real life experiences are highly structured in time, and so to have a complete understanding of the human memory system, our experiments and models of memory must not ignore this fact. Our work provides a theoretical advance in our understanding of what it means to remember as well as a novel methodological tool to study it.

%RESULTS

### 2018-09-02 paper/main.tex
revised commented thoughts
working on discussion [WIP]

The results thus far tell us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories of the episode.  We next carried out a series of analyses aimed at understanding which brain structures might implement these processes.  In one analysis we sought to identify which brain structures were sensitive to the video's dynamic content, as characterized by its topic trajectory.  Specifically, we used a searchlight procedure to identify the extent to which each cluster of voxels exhibited a timecourse (as the participants watched the video) whose temporal correlation matrix matched the temporal correlation matrix of the original video's topic proportion matrix (Fig.~\ref{fig:model}B).  As shown in Figure~\ref{fig:brainz}A, the analysis revealed a network of regions including bilateral frontal and cingulate cortex, suggesting that these regions may play a role in maintaining information relevant to the narrative structure of the video.  In a second analysis, we sought to identify which brain structures' responses (while viewing the video) reflected how each participant would later \textit{recall} the video.  Specifically, we used an analogous searchlight procedure to identify clusters of voxels whose temporal correlation matrices reflected the temporal correlation matrix of the topic proportions for each individual's recalls (Figs.~\corrmats).  As shown in Figure~\ref{fig:brainz}B, the analysis revealed a network of regions including the ventromedial prefrontal cortex, anterior cingulate, and right medial temporal lobe, suggesting that these regions may play a role in transforming each individual's experience into memory.

@@

\section*{Discussion}
\label{sec:discussion}
% JRM THOUGHTS ON DISCUSSION:
% - Nature says "2-3 paragraphs" -- needs to be much shorter.
% - I think we can get rid of anything we mention in the intro or results
% - Things not easily presented elsewhere are the background on the brain results and applications to education metrics.  I think we can focus on those two parts, add an "overview" and "conclusion" (potentially), and cut the rest.
Studying human memory is commonly distilled down to a process of matching specific moments of a past experience with specific mnemonic outcomes.  In traditional trial-based free recall experiments, individual stimuli encountered during encoding are typically labeled as ``remembered'' or ``forgotten'' depending on whether the stimulus is recalled/recognized during a subsequent test. While this approach has advanced our understanding of human memory immensely, it does not translate well to naturalistic experiences. For one thing, the contents of our recall at any given moment might be reflected in many prior experiences/moments. Furthermore, the particular words used to describe the experience will inevitably vary across people and even across repeated recollections within an individual. Thus, there is not a ``one-to-one'' mapping between naturalistic experiences and their mnemonic counterparts. For example, remembering the patterns and colors of each person's shirt in a crowd might be considered as excellent recall in a standard memory task setup. But if the rememberer failed to note that the people in the crowd were gathered for their surprise birthday party, then they would have missed the "point" of the experience.

Our topic modeling approach, whereby we consider the broad ``theme'' present in different moments of participants' experiences and their memories for those experiences, affords us the ability to flexibly and accurately characterize memory for naturalistic experiences. This approach allows us to quantify which moments from the past and the current recounted experience match in terms of their thematic content and critically, our ability to perform this matching does not require participants to use any specific overlapping words. Our work characterizes recollection of an experience by comparing the overall ``shape'' of a dynamic stimulus and a memory. We assess the quality of memory for the video participants viewed by measuring the match between the shapes of the video's trajectory and each participant's recall trajectory. By contrast, the number of recalls could be captured by the "sampling frequency" along that trajectory -- but the number of recalls alone cannot tell us whether participants successfully recollected the meaning of the story by capturing the salient points of the narrative that define its main shape. In addition to providing a way to capture the shape of an experience, this method affords the ability to quantify the particular contents of memory.  Whereas traditional approaches abstract over the content (e.g., percent correct), Our approach allows one to quantify which aspects of an experience are memorable and which fail to stick. This aspect of our approach opens the door for a much richer characterization of memory that considers not just how much information was recalled, but the particular details of that information as well.

Prior work on neural responses during naturalistic experiences and recall has largely focused on identifying brain regions whose responses are reliably similar across individuals ~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17}. This allows one to identify which regions might be processing or representing the stimulus, or retrieving details of the experience during recall. Our approach is fundamentally different. We ask: which brain regions during video viewing match the video structure and how individual participants will recall the video later (in terms of the temporal correlations of their recall topic trajectory)? This latter approach highlights regions such as the medial prefrontal cortex (mPFC) and medial temporal lobes (MTL) that may respond in idiosyncratic ways across individuals, but that nonetheless play an important role in encoding experience into memory. In other work, the mPFC has been suggested to play a role in the representation of a ``schema'', or a network of prior knowledge that is consistent with the current task ~\citep{KestEtal12, GilbMarl17}. One interpretation for mPFC's involvement in our work is that mPFC represents participant-specific relationships with the content of the video, and its representation during scene viewing is related to how a participant later describes the scene. MTL activity patterns also matched participant's recall, which is consistent with a large body of literature highlighting the role of the MTL in episodic memory encoding ~\citep{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.

More broadly, these findings have strong implications for how we assess memory in other naturalistic contexts, such as the classroom or in a doctor's office.  Whereas academic tests often measure students' performance using metrics such as the proportion of correctly answered questions, our work suggests that this approach might ``miss the forrest for the trees''. We view ``true learning'' as understanding key concepts (i.e., understanding central themes in the learned content and how they relate) rather than regurgitating the greatest number of facts. In addition to educational contexts, our approach may provide unique metrics that can be used to assist in the diagnosis of memory disorders, and other psychiatric disorders that influence memory. For example, while the quantity of information recalled could be roughly matched between a healthy and patient population, and other aspects of the memory (such as the shape, serial order, precision or distinctiveness) might be different. Thus, this work serves as a foundation for more nuanced approaches to memory assessment that consider the trajectory and specific contents of memory for a naturalistic experience.

An important question for future work concerns the factors that drive an individual to sample their recall trajectory finely or coarsely. For example, given a short recall interval, would participants intuitively gravitate towards coarser samplings that still outline the basic shape of the video's topic trajectory? Or if participants were told that their narrations would be played back to other participants ~\citep{ZadbEtal17}, would that change the resolution or shape of their recalls? And over successive recountings of the same sequence of events or with more elapsed time between encoding and recall, how do the shapes of the trajectories change? For example, loss of detail would result in a "smoothing out" of the trajectories with each new retelling.

While we view this work as a major advance in characterizing and understanding human memory, as with any approach there are limitations. First, the approach relies on having a ``good'' model of the stimulus (e.g., one that describes the video in the same way as participants recall it), which is currently only achievable by a human hand-annotating each moment of the video.  To increase the scalability of this approach, future work could explore automated methods for extracting meaning from videos ~\citep{YuEtal16}.  Another potential limitation is that by its nature, the model extracts the ``gist'' of scenes from the video.  This provides desirable flexibility (e.g., participants can use different combinations of words to describe the scenes), but this comes at the expense of capturing specific details.  Thus, a future direction of this work will be to increase sensitivity to such details while maintaining flexibility.

To conclude, we'll revisit the question of what it means to remember something. Our view is that ``successful remembering'' is about recovering the ``trajectory'' of an experience, rather than the ability to recognize/recall any of its particular isolated details. Decades of research suggest that episodic memories are not veridical and context-free snapshots of the past, and so treating (and modelling) them as such is overly simplistic at best (and counterproductive at worst). While it's undeniable that these models have been useful in advancing our understanding of human memory, they are severely limited in their ability to explain memory for real life experiences. Real life experiences are highly structured in time, and so to have a complete understanding of the human memory system, our experiments and models of memory must not ignore this fact. Our work provides a theoretical advance in our understanding of what it means to remember as well as a novel methodological tool to study it.

>>>>

The results thus far tell us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories of the episode.  We next carried out a series of analyses aimed at understanding which brain structures might implement these processes.  In one analysis we sought to identify which brain structures were sensitive to the video's dynamic content, as characterized by its topic trajectory.  Specifically, we used a searchlight procedure to identify the extent to which each cluster of voxels exhibited a timecourse (as the participants watched the video) whose temporal correlation matrix matched the temporal correlation matrix of the original video's topic proportion matrix (Fig.~\ref{fig:model}B).  As shown in Figure~\ref{fig:brainz}A, the analysis revealed a network of regions including bilateral frontal and cingulate cortex, suggesting that these regions may play a role in maintaining information relevant to the narrative structure of the video.  In a second analysis, we sought to identify which brain structures' responses (while viewing the video) reflected how each participant would later \textit{recall} the video.  Specifically, we used an analogous searchlight procedure to identify clusters of voxels whose temporal correlation matrices reflected the temporal correlation matrix of the topic proportions for each individual's recalls (Figs.~\corrmats).  As shown in Figure~\ref{fig:brainz}B, the analysis revealed a network of regions including the ventromedial prefrontal cortex (vmPFC), anterior cingulate, and right medial temporal lobe (rMTL), suggesting that these regions may play a role in transforming each individual's experience into memory.

@@

\section*{Discussion}
\label{sec:discussion}

One approach to identifying neural responses to naturalistic stimuli (including experiences) entails building a model of the stimulus and searching for brain regions whose responses are consistent with the model.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have extended this approach with a clever twist.  Rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time.  By contrast, in our present work we used topic models and HMMs to construct an explicit stimulus model (i.e., the topic trajectory of the movie).  When we searched for brain structures whose responses are consistent with the video's topic trajectory, we identified a network of structures that overlaped strongly with the ``long temporal receptive window'' network reported by the Hasson group~\citep[e.g., compare our Fig.~\ref{fig:brainz}A with the map of long temporal receptive window voxels in ][]{LernEtal11}.  This provides support for the notion that part of the long temporal receptive window network may be maintaining an explicit model of the stimulus dynamics.  When we performed a similar analysis after swapping out the video's topic trajectory with the recall topic trajectories of each individual participant, this allowed us to identify brain regions whose responses (as the participants viewed the video) reflected how the video trajectory would be transformed in memory (as reflected by the recall topic trajectory).  The analysis revealed that the rMTL and vmPFC may play a role in this person-specific transformation from experience into memory.  The role of the MTL in episodic memory encoding has been well-reported~\citep[e.g., ][]{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.  Prior work has also implicated the medial prefrontal cortex in representing ``schema'' knowledge~\citep[i.e., general knowledge about the format of an ongoing experience given prior similar experiencs; ][]{KestEtal12, GilbMarl17}.  Integrating across our study and this prior work, one interpretation is that the person-specific transformations mediated (or represented) by the rMTL and vmPFC may reflect schema knowledge being leveraged, formed, or updated, incorporating ongoing experience into previously acquired knowledge.

Our work has broad implications for how we characterize and assess memory in real-world settings such as the classroom or physician's office.  For example, the most commonly used classroom evaluation tools involve computing the proportion of correctly answered exam questions.  Our work indicates that this approach is only loosly related to what educators might really want to measure: how well did the students understand the key ideas presented in the course?  One could apply the computational framework we developed to construct topic trajectories for the video and participants' recalls to build explicit content models of the course material and exam questions.  This approach would provide a more nuanced and specific view into which aspects of the material students had learned well (or poorly).  Similarly, memory evaluations that incorporate explicit content models might provide more direct measures of the qualitiy of patients' memories.

% JRM THOUGHTS ON DISCUSSION:
% - Nature says "1-2 short paragraphs" -- needs to be much shorter.
% - I think we can get rid of anything we mention in the intro or results
% - Things not easily presented elsewhere are the background on the brain results and applications to education metrics.  I think we can focus on those two parts, add an "overview" and "conclusion" (potentially), and cut the rest.
